\batcave: Calling somatic mutations with a tumor- and site-specific prior

### 2019-12-09 NAR_DRAFT.tex
small changes to copy number caveat

\batcave divides the data into two classes: high- and low-confidence variants.
The high-confidence variants are used to estimate the mutational prior and mutation rate, which are then used to improve the calling of low-confidence variants.
Statistically, this is an empirical Bayesian approach~\cite{Robbins1954}, in which the high and low-confidence variants are treated as parallel experiments \cite{Morris1983,Efron2014}. 
In general, high-confidence variants tend to have relatively high allelic frequencies, and consequently tend to have arisen early in tumor development.
An  implicit assumption of our approach is that the mutational process does not change between high- and low-confidence variants, implying that the mutational profile of the tumor is temporally constant.
Recent studies have found differences in mutational profiles among variants of different allelic frequencies~\cite{Rubanova2018a}, and although in general those differences are relatively small, they can be significant in some cases.
\bkmcomment{We provide the user with the ability to output a file with the contents of the estimated mutation profile as each high-confidence variant is evaluated.
In cases where the user suspects a shift in mutational process, we encourage using this file to verify that the final mutation profile is representative of the processes active at low frequency.}
A potential extension of the \batcave algorithm is to process potential variants in order of descending allelic frequency and to update the estimated mutational prior as the algorithm proceeds.
This approach might increase sensitivity to low-frequency variants generated by recently-arisen mutational processes, at the cost of potentially increasing sensitivity to patterns of sequencing error.
\bkmcomment{We use properties of the allele frequency distribution generated by mutation acquisition during exponential growth.
Copy number alterations can distort the allele frequency distribution.
While copy number alterations in a large number of genomic regions would be required to significantly affect the estimate of mutation rate, an additional extension of the \batcave algorithm could be to incorporate external copy number calls in order to adjust allele frequencies.}
Our results show that adding a mutational prior substantially improves probabilistic variant calling, particularly for tumors with concentrated profiles.
Improved variant calling increases the benefit-to-cost ratio of deep sequencing in both research and clinical applications.
Moreover, \batcave proves to be a better calibrated caller than vanilla MuTect (Fig.~\ref{NAR-ppv_fig}).
Different users will prefer different tradeoffs in terms of precision and recall, which can be more accurately made with \batcave.
Our R implementation, \texttt{batcaver}, can be easily incorporated into any MuTect-based pipeline, and the mutational profile algorithm can be incorporated into many other callers.

>>>>

\batcave divides the data into two classes: high- and low-confidence variants.
The high-confidence variants are used to estimate the mutational prior and mutation rate, which are then used to improve the calling of low-confidence variants.
Statistically, this is an empirical Bayesian approach~\cite{Robbins1954}, in which the high and low-confidence variants are treated as parallel experiments \cite{Morris1983,Efron2014}. 
In general, high-confidence variants tend to have relatively high allelic frequencies, and consequently tend to have arisen early in tumor development.
An  implicit assumption of our approach is that the mutational process does not change between high- and low-confidence variants, implying that the mutational profile of the tumor is temporally constant.
Recent studies have found differences in mutational profiles among variants of different allelic frequencies~\cite{Rubanova2018a}, and although in general those differences are relatively small, they can be significant in some cases.
\bkmcomment{We provide the user with the ability to output a file with the contents of the estimated mutation profile as each high-confidence variant is evaluated.
In cases where the user suspects a shift in mutational process, we encourage using this file to verify that the final mutation profile is representative of the processes active at low frequency.}
A potential extension of the \batcave algorithm is to process potential variants in order of descending allelic frequency and to update the estimated mutational prior as the algorithm proceeds.
This approach might increase sensitivity to low-frequency variants generated by recently-arisen mutational processes, at the cost of potentially increasing sensitivity to patterns of sequencing error.
\bkmcomment{We use properties of the allele frequency distribution of high-confidence mutations to estimate the mutation rate in a tumor, and copy number alterations can distort the allele frequency distribution.
While copy number alterations in a large number of genomic regions would be required to significantly affect the estimate of mutation rate, an additional extension of the \batcave algorithm could be to incorporate external copy number calls in order to adjust allele frequencies.}
Our results show that adding a mutational prior substantially improves probabilistic variant calling, particularly for tumors with concentrated profiles.
Improved variant calling increases the benefit-to-cost ratio of deep sequencing in both research and clinical applications.
Moreover, \batcave proves to be a better calibrated caller than vanilla MuTect (Fig.~\ref{NAR-ppv_fig}).
Different users will prefer different tradeoffs in terms of precision and recall, which can be more accurately made with \batcave.
Our R implementation, \texttt{batcaver}, can be easily incorporated into any MuTect-based pipeline, and the mutational profile algorithm can be incorporated into many other callers.
